Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-03-23 analysis/paper/paper.Rmd
Updated with edits from IC and MR

Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` in fall 2019 and from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` in fall 2020 (Table \@ref(tab:abund-tab)). For both surveys, the unadjusted multiple-census estimates were larger than the single census estimates and small numbers of recaptures resulted in wide confidence intervals (Figure \@ref(fig:abund-fig)). Because our sampling design most closely matched a multiple census estimator and required a small sample size modification, we deemed the Schnabel estimates most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_F, spring_abund$Uci_F)), big.mark = ","), collapse = " - ")`). Using the delayed-mixing Schnabel estimator, the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish/100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

@@

We found that gastric lavage successfully removed food items in all dissected individuals, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs. Northern Pikeminnow captured during spring 2021 had a higher proportion of fish content in their diet compared to individuals captured in the fall surveys (Table \@ref(tab:gastric-tab)). The mean wet weight of total contents for an individual was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; sd = `r round(sd(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants made up `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of all stomach contents examined.

@@

During the fall DSR emigration, we estimate an average-size Northern Pikeminnow (`r round(avg_fall_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"]` g of fish to maintain their body size, assuming 60% of their diet consists of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f30_g"]` g for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f90_g"]` g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_spring_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"]` g of fish, assuming 60% of their diet is fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f30_g"]` for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f90_g"]` for a diet of 90% fish. Given estimated mean Northern Pikeminnow population sizes of `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` from fall estimates and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` for spring 2021, and assuming 60% of their diet is fish, we estimate that the Northern Pikeminnow population can consume `r round(tot_g_DSR, 1)` metric tons of fish during the fall DSR emigration and an additional `r round(tot_g_NRR, 1)` metric tons during the spring NRR emigration in Deadwater Slough.

@@

Using the total fish consumed by the Northern Pikeminnow population (summing fall and spring surveys), we produced a sensitivity analysis for three Northern Pikeminnow diet scenarios where 60% of the diet is fish, and juvenile Chinook Salmon make up 30%, 50%, and 65% of total fish prey consumed (Figure \@ref(fig:juv-con-p)). Assuming that 50% of the total fish consumed are juvenile Chinook salmon, we estimate that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$juv_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$juv_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) juvenile Chinook Salmon will be consumed. Using the median Granite-to-Granite SAR, we estimated that juvenile consumption is equivalent to `r prettyNum(round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$adult_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$adult_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) adults (Figure \@ref(fig:adult-imp-p)).

@@

Spawner abundance is perhaps the most important metric considered in determining a population's viability and productivity. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces natural-origin returns to upriver populations by `r round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50])` adults, annually, which is `r round((total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50] / 558) * 100)`% of the total mean adult escapement estimated<!--of 558--> for the Upper Salmon MPG in 2017 - 2019 [@Kinzer2020]. Even under the most conservative scenario where 30% of the Northern Pikeminnow diet is fish and 30% of that is Chinook Salmon we estimate the "adult equivalents" of the juveniles consumed to be `r round(total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30])`, which is `r round((total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30] / 558) * 100)`% of recent natural-origin adult escapements. In addition to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon, including hatchery releases, are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that adult returns of steelhead and Sockeye Salmon are also affected by predation, including hatchery populations, which provide recreational fishing opportunities. Consequently, reducing predation mortality at Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including other ESA-listed species. Because the Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population.

>>>>

Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` in fall 2019 and from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` in fall 2020 (Table \@ref(tab:abund-tab)). We estimated larger populations of Northern Pikeminnow using the unadjusted multiple-census estimator compared to adjusted and single census estimators (Figure \@ref(fig:abund-fig)). Because our sampling design most closely matched a multiple census estimator and required a small sample size modification, we deemed the Schnabel estimates most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_F, spring_abund$Uci_F)), big.mark = ","), collapse = " - ")`). Using the delayed-mixing Schnabel estimator, the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish/100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

@@

We found that gastric lavage successfully removed food items in all dissected individuals, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs. Northern Pikeminnow captured during spring 2021 had a higher proportion of fish content in their diet compared to individuals captured in the fall surveys (Table \@ref(tab:gastric-tab)). The mean wet weight of total contents for an individual was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; SD = `r round(sd(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants made up `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of all stomach contents examined.

@@

During the fall DSR emigration, we estimate an average-size Northern Pikeminnow (`r round(avg_fall_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"]` g of fish to maintain their body size, assuming 60% of their diet consists of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f30_g"]` g for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f90_g"]` g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_spring_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"]` g of fish, assuming 60% of their diet is fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f30_g"]` for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f90_g"]` for a diet of 90% fish. Given estimated mean Northern Pikeminnow population sizes of `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` in the fall, `r prettyNum(round(spring_abund$N_S), big.mark = ",")` in the spring 2021, and a diet consisting of 60% fish, the Northern Pikeminnow population can consume `r round(tot_g_DSR, 1)` metric tons of fish during the fall DSR emigration and `r round(tot_g_NRR, 1)` metric tons during the spring NRR emigration in Deadwater Slough.

@@

Given yearly fish consumption estimates during the fall DSR and spring NRR periods, we produced a sensitivity analysis to calculated consumption of juvenile Chinook Salmon where 30%, 50%, and 65% of total fish prey consumed are juvenile Chinook Salmon (Figure \@ref(fig:juv-con-p)). At the median value of 50%, we estimated that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$juv_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$juv_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) juvenile Chinook Salmon will be consumed. Using the median Granite-to-Granite SAR, we estimated that juvenile consumption is equivalent to `r prettyNum(round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$adult_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$adult_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) adults (Figure \@ref(fig:adult-imp-p)).

@@

Spawner abundance is perhaps the most important metric considered in determining a population's viability and productivity. We estimated that consumption of juvenile Chinook Salmon by Northern Pikeminnow in the Deadwater Slough potentially reduces natural-origin returns to upriver populations by `r round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50])` adults, annually, which is `r round((total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50] / 558) * 100)`% of the total mean adult escapement estimated<!--of 558--> for the Upper Salmon MPG in 2017 - 2019 [@Kinzer2020]. Even under the most conservative scenario where 30% of the Northern Pikeminnow diet is fish and 30% of that is juvenile Chinook Salmon, we estimated `r round(total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30])` "adult equivalents" would be consumed, which is approximately `r round((total_impacts$adult_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30] / 558) * 100)`% of recent natural-origin adult Chinook Salmon escapements. In addition to juvenile Chinook Salmon, juvenile steelhead and Sockeye Salmon, including hatchery releases, are also likely prey items for Northern Pikeminnow in Deadwater Slough. Accordingly, we surmise that adult returns of steelhead and Sockeye Salmon are also affected by predation, including hatchery populations, which provide recreational fishing opportunities. Consequently, reducing predation mortality at Deadwater Slough could potentially benefit multiple upriver natural and hatchery populations, including other ESA-listed species. Because the Deadwater Slough is part of the migratory pathway for multiple species and populations of emigrating salmonids, the positive impact could be greater than individual tributary rehabilitation actions which typically benefit a single population.
